---
title: "Test functions"
format: html
editor: visual
author: "Marcelo Hurtado"
---

## Setup

```{r setup}
library(dplyr)
library(BioinfoTools)
set.seed(123)
dir.create("Results", showWarnings = FALSE)
```

## Data example

```{r}
counts <- as.matrix(read.csv("~/Documents/MIRACLE/input/raw_files/Vanderbilt/Counts_Vanderbilt_SYMBOL.csv", row.names = 1))
clinical <- read.csv("~/Documents/MIRACLE/input/raw_files/Vanderbilt/ClinicalData_Vanderbilt.csv", row.names = 1)
survival <- read.csv("~/Documents/LungPredict1_paper_complete/Input/Survival_data_Vanderbilt.csv", row.names = 1)

clinical <- clinical[colnames(counts), ]
clinical$PFS <- survival[as.character(clinical$pt_ID), "PFS"]
clinical$PFS_event <- ifelse(survival[as.character(clinical$pt_ID), "DRP_st"] == "Yes", 1, 0)
```

Log-TPM normalization via `compute.normalization()` (`zscore = FALSE`, since the TF and pathway activity
functions expect normalized expression, not per-gene standardized values), used as input for those functions.

```{r}
counts.norm <- as.matrix(compute.normalization(counts, zscore = FALSE))
```

## `compute.TFs.activity()`

Infers TF activity from normalized expression (VIPER) using a TF-target regulon.

CollecTRI network (default):

```{r}
tfs_collectri <- compute.TFs.activity(counts.norm, file.name = "CollecTRI")
head(tfs_collectri[, 1:5])
```

Dorothea network. `Results/TF_target_collection.csv` caches whichever collection was fetched last regardless of `TF.collection`, so it is removed first to force a genuine Dorothea fetch:

```{r}
file.remove("Results/TF_target_collection.csv")
tfs_dorothea <- compute.TFs.activity(counts.norm, TF.collection = "Dorothea", file.name = "Dorothea")
head(tfs_dorothea[, 1:5])
```

ARACNE network, from a user-supplied `network.txt` path (a LUAD network built for this cohort):

```{r}
tfs_aracne <- compute.TFs.activity(counts.norm, TF.collection = "ARACNE",
                                    aracne.network = "~/Documents/ARACNE/luad/network/network.txt",
                                    file.name = "ARACNE")
head(tfs_aracne[, 1:5])
```

A single underlying `decoupleR` method instead of the default `"consensus"` ensemble:

```{r}
tfs_mlm <- compute.TFs.activity(counts.norm, statistic = "mlm", file.name = "CollecTRI_mlm")
head(tfs_mlm[, 1:5])
```

Every `statistic` option, run on the (cached) CollecTRI network. A small sample subset keeps the slower
tree-based methods (`udt`, `mdt`) fast; `"gsva"` is skipped since `decoupleR`'s `run_gsva()` is incompatible
with the installed GSVA \>= 1.50 API:

```{r}
all_statistics <- c("consensus", "aucell", "udt", "mdt", "wmean", "ulm", "mlm", "wsum", "viper", "ora", "fgsea")
counts.norm.small <- counts.norm[, 1:15]

for (st in all_statistics) {
  tfs_st <- compute.TFs.activity(counts.norm.small, statistic = st, file.name = paste0("CollecTRI_", st))
  cat(st, "-> dim:", dim(tfs_st), "\n")
}
```

## `compute.pathway.activity()`

Computes pathway activity from `pathway_source` (PROGENy by default, or KEGG/REACTOME/Hallmark), scored by
any `decoupleR` `statistic` -- the same pattern as `compute.TFs.activity()`. `gene_sets`, if provided, is
scored with that same `statistic` and saved to `Results/`, but the function always returns only the
`pathway_source` matrix.

PROGENy only (default `pathway_source`/`statistic`):

```{r}
pathways_progeny <- compute.pathway.activity(counts.norm, file.name = "PROGENy")
head(pathways_progeny[, 1:5])
```

PROGENy + custom gene sets (MSigDB Hallmark, scored with the same default `"consensus"` statistic and
saved to `Results/Pathway_matrix_gene_sets_Hallmark.csv`, but not part of the return value):

```{r}
hallmark <- msigdbr::msigdbr(species = "Homo sapiens", category = "H")
gene_sets <- split(hallmark$gene_symbol, hallmark$gs_name)
pathways_hallmark <- compute.pathway.activity(counts.norm, gene_sets = gene_sets, file.name = "Hallmark")
head(pathways_hallmark[, 1:5])
```

`pathway_source` swaps PROGENy for a built-in MSigDB collection (KEGG, REACTOME, or Hallmark):

```{r}
pathways_kegg <- compute.pathway.activity(counts.norm, pathway_source = "KEGG", file.name = "KEGG")
head(pathways_kegg[, 1:5])
```

A single underlying `decoupleR` method instead of the default `"consensus"` ensemble:

```{r}
pathways_kegg_viper <- compute.pathway.activity(counts.norm, pathway_source = "KEGG", statistic = "viper", file.name = "KEGG_viper")
head(pathways_kegg_viper[, 1:5])
```

`"mlm"` (and so the default `"consensus"`) fits all sources jointly, which can fail on large, highly
redundant collections like REACTOME's 1615 pathways -- a per-source method such as `"ulm"` still works:

```{r}
tryCatch(
  compute.pathway.activity(counts.norm, pathway_source = "REACTOME", statistic = "mlm", file.name = "REACTOME_mlm"),
  error = function(e) cat("Expected error:", conditionMessage(e), "\n")
)
pathways_reactome_ulm <- compute.pathway.activity(counts.norm, pathway_source = "REACTOME", statistic = "ulm", file.name = "REACTOME_ulm")
dim(pathways_reactome_ulm)
```

## `run_deg_analysis()`

Differential expression (edgeR/limma-voom) between CANARY radiologic risk classes. A volcano plot (built
from the full, unfiltered gene set) is saved to `Results/` by default (`plot = TRUE`).

Default factor ordering (Intermediate vs. Poor):

```{r}
sub <- clinical$CANARY %in% c("I", "P")
deg_default <- run_deg_analysis(counts[, sub], clinical[sub, ], group_col = "CANARY", file.name = "CANARY_IvsP")
head(deg_default)
```

Explicit reference level (Good vs. Intermediate, `ref_level = "I"`):

```{r}
sub2 <- clinical$CANARY %in% c("G", "I")
deg_ref <- run_deg_analysis(counts[, sub2], clinical[sub2, ], group_col = "CANARY", ref_level = "I", file.name = "CANARY_GvsI")
head(deg_ref)
```

## `compute.volcano.plot()`

Called automatically by `run_deg_analysis()` above, but also usable standalone on any DEG result. Here on
the full, unfiltered gene set (`pval = 1`), with a stricter `logFC` cutoff:

```{r}
deg_full <- run_deg_analysis(counts[, sub2], clinical[sub2, ], group_col = "CANARY", ref_level = "I", pval = 1, plot = FALSE)
compute.volcano.plot(deg_full, pval = 0.05, logFC = 1.5, title = "CANARY: G vs. I", file.name = "CANARY_GvsI_standalone")
```

## `compute.enrichment()`

Over-representation (ORA, default `method`) or gene set enrichment (GSEA) against GO (BP/CC/MF), KEGG,
REACTOME, PROGENy, or Hallmark; a dot plot is saved per call to `Results/`.

ORA on the significant genes from `deg_ref` (CANARY G vs. I), single database:

```{r}
ora_kegg <- compute.enrichment(rownames(deg_ref), database = "KEGG", universe = rownames(counts[, sub2]), file.name = "KEGG")
head(ora_kegg@result[, c("Description", "pvalue", "p.adjust")])
```

Every `database` option for ORA:

```{r}
databases <- c("GO_BP", "GO_CC", "GO_MF", "KEGG", "REACTOME", "PROGENy", "Hallmark")

for (db in databases) {
  ora_db <- compute.enrichment(rownames(deg_ref), database = db, universe = rownames(counts[, sub2]), file.name = paste0("ORA_", db))
  cat(db, "-> n significant terms:", if (is.null(ora_db)) 0 else nrow(ora_db@result), "\n")
}
```

`method = "GSEA"` needs a full ranked gene list, not pre-filtered to significant genes
(`run_deg_analysis(..., pval = 1)`):

```{r}
deg_full <- run_deg_analysis(counts[, sub2], clinical[sub2, ], group_col = "CANARY", ref_level = "I", pval = 1, plot = FALSE)
ranks <- setNames(deg_full$t, rownames(deg_full))

gsea_hallmark <- compute.enrichment(ranks, database = "Hallmark", method = "GSEA", file.name = "Hallmark")
head(gsea_hallmark@result[, c("Description", "NES", "p.adjust")])
```

Every `database` option for GSEA (REACTOME is the slow one here: ~1615 pathways against the full ranked list):

```{r}
for (db in databases) {
  gsea_db <- compute.enrichment(ranks, database = db, method = "GSEA", file.name = paste0("GSEA_", db))
  cat(db, "-> n significant terms:", if (is.null(gsea_db)) 0 else nrow(gsea_db@result), "\n")
}
```

## `compute.TFs.activity.DEG()`

TF activity computed directly from a per-gene DEG ranking statistic (a named vector the caller builds
and chooses -- the `t` statistic is recommended) instead of per-sample expression, with an empirical
p-value from gene-label permutations in addition to the score. Default `statistic` is `"consensus"`,
same as `compute.TFs.activity()`:

```{r}
signature <- setNames(deg_full$t, rownames(deg_full))
tf_deg <- compute.TFs.activity.DEG(signature, n_perm = 200, file.name = "CANARY_GvsI")
head(tf_deg$activity)
tf_deg$plot
```

`TF.collection = "ARACNE"` needs the full normalized expression matrix (`RNA.counts`) to derive each
edge's mode of regulation via cross-sample correlation, since that can't come from the single-column
signature being scored:

```{r}
tf_deg_aracne <- compute.TFs.activity.DEG(signature, TF.collection = "ARACNE",
                                          aracne.network = "~/Documents/ARACNE/luad/network/network.txt",
                                          RNA.counts = counts.norm, statistic = "ulm", n_perm = 200, file.name = "ARACNE")
head(tf_deg_aracne$activity)
```

## `compute.survival.analysis()`

Kaplan-Meier analysis and log-rank test; plots are saved to `Results/`.

Predefined clinical grouping (`group_column`):

```{r}
km_group <- compute.survival.analysis(clinical, PFS = "PFS", PFS_event = "PFS_event",
                                       group_column = "CANARY", file_name = "CANARY")
km_group$p_value
km_group$median_PFS
```

Automatic screening across a feature matrix (`features`), High/Low split per PROGENy pathway score:

```{r}
clin_feat <- clinical[rownames(pathways_progeny), ]
km_features <- compute.survival.analysis(clin_feat, PFS = "PFS", PFS_event = "PFS_event",
                                          features = pathways_progeny, p.value = 0.15, file_name = "PROGENy")
names(km_features)
```

## `compute.normalization()`

Raw counts to log-TPM, optionally z-score scaled per gene.

Default (log-TPM + z-score):

```{r}
norm_default <- compute.normalization(counts)
head(norm_default[, 1:5])
```

Without z-score scaling:

```{r}
norm_noz <- compute.normalization(counts, zscore = FALSE)
head(norm_noz[, 1:5])
```

## `compute.pca.analysis()`

PCA on a samples x features score matrix; saves a PC1/PC2 plot to `Results/`.

Colored by a clinical grouping column:

```{r}
pca_group <- compute.pca.analysis(pathways_progeny, coldata = clinical, group_col = "CANARY", file.name = "PROGENy_CANARY")
pca_group$var_explained
```

Without a grouping column:

```{r}
pca_nogroup <- compute.pca.analysis(pathways_progeny, file.name = "PROGENy")
head(pca_nogroup$coords)
```

## `compute.correlation()`

Pearson and Spearman correlation, dispatching on the shape of the inputs.

Two numeric vectors:

```{r}
compute.correlation(clinical$Age_at_collection, clinical$BMI)
```

Score matrix vs. a continuous trait:

```{r}
corr_trait <- compute.correlation(pathways_progeny, coldata = clinical, trait = "BMI", pval = 0.1)
corr_trait$stats
```

Two matrices (also saves a correlation heatmap per method to `Results/`):

```{r}
corr_matrices <- compute.correlation(tfs_collectri[, 1:8], pathways_progeny, file_name = "TF_vs_PROGENy")
corr_matrices$pearson$cor[1:3, 1:3]
```

## `scores.*.test()` family

Each tests a samples x features score matrix directly against a continuous trait, plotting
significant hits to `Results/`.

```{r}
pearson_sig  <- scores.pearson.test(pathways_progeny, clinical, trait = "BMI", pval = 0.1)
spearman_sig <- scores.spearman.test(pathways_progeny, clinical, trait = "BMI", pval = 0.1)
fisher_sig   <- scores.fisher.test.continuous(pathways_progeny, clinical, trait = "BMI", pval = 0.3)

colnames(pearson_sig)
```

## `scores.stat.analysis.continuous()`

Unified dispatcher for the `scores.*.test()` family above, selected via `method`.

```{r}
wrap_spearman <- scores.stat.analysis.continuous(pathways_progeny, clinical, trait = "BMI", method = "spearman", pval = 0.1)
colnames(wrap_spearman)
```

## Categorical `scores.*.test()` family

Each tests a samples x features score matrix against a categorical trait, plotting significant
hits (boxplot + jitter) to `Results/`. `ttest`/`wilcox` require exactly 2 groups; `anova`/`kruskal`
accept 2+.

```{r}
sub <- clinical$CANARY %in% c("I", "P")
ttest_sig  <- scores.ttest.test(pathways_progeny[sub, ], clinical[sub, ], trait = "CANARY", pval = 0.5)
wilcox_sig <- scores.wilcox.test(pathways_progeny[sub, ], clinical[sub, ], trait = "CANARY", pval = 0.5)
anova_sig  <- scores.anova.test(pathways_progeny, clinical, trait = "CANARY", pval = 0.5)
kruskal_sig <- scores.kruskal.test(pathways_progeny, clinical, trait = "CANARY", pval = 0.5)

colnames(anova_sig)
```

## `scores.stat.analysis.categorical()`

Unified dispatcher for the categorical `scores.*.test()` family above, selected via `method`.

```{r}
wrap_kruskal <- scores.stat.analysis.categorical(pathways_progeny, clinical, trait = "CANARY", method = "kruskal", pval = 0.5)
colnames(wrap_kruskal)
```