---
title: "functions.r — tests on the Vanderbilt cohort (LungPredict1 paper)"
format: html
editor: visual
---

## Setup

```{r setup}
library(dplyr)
source("functions.r")
set.seed(123)
dir.create("Results", showWarnings = FALSE)
```

## Data

Vanderbilt early-detection cohort (raw counts, clinical and survival data) from the LungPredict1 paper.

```{r}
counts <- as.matrix(read.csv("~/Documents/MIRACLE/input/raw_files/Vanderbilt/Counts_Vanderbilt_SYMBOL.csv", row.names = 1))
clinical <- read.csv("~/Documents/MIRACLE/input/raw_files/Vanderbilt/ClinicalData_Vanderbilt.csv", row.names = 1)
survival <- read.csv("~/Documents/LungPredict1_paper_complete/Input/Survival_data_Vanderbilt.csv", row.names = 1)

clinical <- clinical[colnames(counts), ]
clinical$PFS <- survival[as.character(clinical$pt_ID), "PFS"]
clinical$PFS_event <- ifelse(survival[as.character(clinical$pt_ID), "DRP_st"] == "Yes", 1, 0)
```

Log2-CPM normalization, used as input for the TF and pathway activity functions.

```{r}
counts.norm <- edgeR::cpm(counts, log = TRUE)
```

## `compute.TFs.activity()`

Infers TF activity from normalized expression (VIPER) using a TF-target regulon.

CollecTRI network (default):

```{r}
tfs_collectri <- compute.TFs.activity(counts.norm, file.name = "CollecTRI")
head(tfs_collectri[, 1:5])
```

Dorothea network. `Results/TF_target_collection.csv` caches whichever collection was fetched last regardless of `TF.collection`, so it is removed first to force a genuine Dorothea fetch:

```{r}
file.remove("Results/TF_target_collection.csv")
tfs_dorothea <- compute.TFs.activity(counts.norm, TF.collection = "Dorothea", file.name = "Dorothea")
head(tfs_dorothea[, 1:5])
```

ARACNE network (cancer-type-specific, LUAD):

```{r}
tfs_aracne <- compute.TFs.activity(counts.norm, TF.collection = "ARACNE", cancer.type = "luad", file.name = "ARACNE")
head(tfs_aracne[, 1:5])
```

## `compute.pathway.activity()`

Computes PROGENy pathway activity scores, optionally combined with GSVA on custom gene sets.

PROGENy only:

```{r}
pathways_progeny <- compute.pathway.activity(counts.norm, file.name = "PROGENy")
head(pathways_progeny[, 1:5])
```

PROGENy + GSVA (MSigDB Hallmark gene sets):

```{r}
hallmark <- msigdbr::msigdbr(species = "Homo sapiens", category = "H")
gene_sets <- split(hallmark$gene_symbol, hallmark$gs_name)
pathways_hallmark <- compute.pathway.activity(counts.norm, gene_sets = gene_sets, file.name = "Hallmark")
head(pathways_hallmark$PROGENy[, 1:5])
head(pathways_hallmark$GSVA[, 1:5])
```

## `run_deg_analysis()`

Differential expression (edgeR/limma-voom) between CANARY radiologic risk classes.

Default factor ordering (Intermediate vs. Poor):

```{r}
sub <- clinical$CANARY %in% c("I", "P")
deg_default <- run_deg_analysis(counts[, sub], clinical[sub, ], group_col = "CANARY")
head(deg_default)
```

Explicit reference level (Good vs. Intermediate, `ref_level = "I"`):

```{r}
sub2 <- clinical$CANARY %in% c("G", "I")
deg_ref <- run_deg_analysis(counts[, sub2], clinical[sub2, ], group_col = "CANARY", ref_level = "I")
head(deg_ref)
```

## `compute.survival.analysis()`

Kaplan-Meier analysis and log-rank test; plots are saved to `Results/`.

Predefined clinical grouping (`group_column`):

```{r}
km_group <- compute.survival.analysis(clinical, PFS = "PFS", PFS_event = "PFS_event",
                                       group_column = "CANARY", file_name = "CANARY")
km_group$p_value
km_group$median_PFS
```

Automatic screening across a feature matrix (`features`), High/Low split per PROGENy pathway score:

```{r}
clin_feat <- clinical[rownames(pathways_progeny), ]
km_features <- compute.survival.analysis(clin_feat, PFS = "PFS", PFS_event = "PFS_event",
                                          features = pathways_progeny, p.value = 0.1, file_name = "PROGENy")
names(km_features)
```

## `compute.normalization()`

Raw counts to log-TPM, optionally z-score scaled per gene.

Default (log-TPM + z-score):

```{r}
norm_default <- compute.normalization(counts)
head(norm_default[, 1:5])
```

Without z-score scaling:

```{r}
norm_noz <- compute.normalization(counts, zscore = FALSE)
head(norm_noz[, 1:5])
```

## `compute.pca.analysis()`

PCA on a samples x features score matrix; saves a PC1/PC2 plot to `Results/`.

Colored by a clinical grouping column:

```{r}
pca_group <- compute.pca.analysis(pathways_progeny, coldata = clinical, group_col = "CANARY", file.name = "PROGENy_CANARY")
pca_group$var_explained
```

Without a grouping column:

```{r}
pca_nogroup <- compute.pca.analysis(pathways_progeny, file.name = "PROGENy")
head(pca_nogroup$coords)
```

## `compute.correlation()`

Pearson and Spearman correlation, dispatching on the shape of the inputs.

Two numeric vectors:

```{r}
compute.correlation(clinical$Age_at_collection, clinical$BMI)
```

Score matrix vs. a continuous trait:

```{r}
corr_trait <- compute.correlation(pathways_progeny, coldata = clinical, trait = "BMI", pval = 0.1)
corr_trait$stats
```

Two matrices (also saves a correlation heatmap per method to `Results/`):

```{r}
corr_matrices <- compute.correlation(tfs_collectri[, 1:8], pathways_progeny, file_name = "TF_vs_PROGENy")
corr_matrices$pearson$cor[1:3, 1:3]
```

## `scores.*.test()` family

Each tests a samples x features score matrix (wrapped in a list, e.g. `list(pathways_progeny)`) against a continuous trait, plotting significant hits to `Results/`.

```{r}
scores_list <- list(pathways_progeny)

pearson_sig  <- scores.pearson.test(scores_list, clinical, trait = "BMI", pval = 0.1)
spearman_sig <- scores.spearman.test(scores_list, clinical, trait = "BMI", pval = 0.1)
kendall_sig  <- scores.kendall.test(scores_list, clinical, trait = "BMI", pval = 0.1)
lm_sig       <- scores.lm.test(scores_list, clinical, trait = "BMI", pval = 0.1)
fisher_sig   <- scores.fisher.test.continuous(scores_list, clinical, trait = "BMI", pval = 0.3)

colnames(pearson_sig[[1]])
```

## `scores.stat.analysis.continuous()`

Unified dispatcher for the `scores.*.test()` family above, selected via `method`.

```{r}
wrap_kendall <- scores.stat.analysis.continuous(pathways_progeny, clinical, trait = "BMI", method = "kendall", pval = 0.1)
colnames(wrap_kendall[[1]])
```

Also accepts NMF-style output with a `$Z` score matrix (e.g. `compute.latent_factors()`):

```{r}
wrap_nmf <- scores.stat.analysis.continuous(list(Z = pathways_progeny), clinical, trait = "BMI", method = "spearman", pval = 0.1)
colnames(wrap_nmf[[1]])
```
